\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-15 k-planar.tex
Arxiv v1

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

\usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

\setlength{\parskip}{1ex}

% \newcommand{\note}[2]{{\color{red}[#1:~#2]}}
% \newcommand{\notex}[2]{{\color{blue}[#1:~#2]}}
\newcommand{\notex}[2]{}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centred colouring, and implicit graph encoding. We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

% Layered partitions of graphs (with small layered width, in which the quotient has small treewidth) are a recently-introduced tool that have been used to solve longstanding problems on queue layouts, non-repetitive colouring, and 3-d graph drawing.  Such partitions are known to exist for planar graphs and, more generally, apex-minor-free graphs.  In the current paper, we prove that every $k$-planar graph has a layered partition of layered width $O(k^2)$ in  which the quotient graph has treewidth $O(k^3)$. This is the first result of this type for a non-minor-closed class of graphs and implies that $k$-planar graphs have both queue number and non-repetitive chromatic number upper-bounded by a function of $k$. The former result was previously shown by combining layered partitions of planar graphs with \textit{ad hoc} methods. The latter result is new, for all $k\ge 1$.  These results extend to $(g,k)$-planar graphs, the natural generalization of $k$-planar graphs to to genus-$g$ surfaces (rather than the genus-$0$ plane).

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002), and \citet{bonamy.gavoille.ea:shorter} use it to find shorter implicit representations of planar graphs (making progress on a sequence of results going back to 1988 \citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}).

@@

Although $k$-planar graphs are the most high-profile target for a generalization \cref{PlanarBasic}, we actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

This definition is related to $k$-planarity because of the following observation

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is isomorphic to a subgraph of $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}, 
% and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

%\note{PM}{TODO: Write here about applications. non-repetitive colouring, $p$-centered colouring, $(g,k)$-planar graph, $(g,d)$-map graphs, $k$-NN graphs, David's rough characterization.}

@@

We now formally define $k$-planar graphs.  An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.  Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

% \note{DW}{I  prefer to not define ``layered $H$-partition''. Just define the ``layered width'' of an $H$-partition, which is what we did in \citep{dujmovic.joret.ea:planar}. There are many places where we write ``$H$-partition with layered width'' without writing ``layered $H$-partition''. I prefer it this way.}

@@

\begin{lem}[\citep{dujmovic.joret.ea:planar}] 
\label{MakeWidth1}
If a graph $G$ has an $H$-partition of layered width $\ell$ with respect to layering $\langle V_0,V_1,\dots\rangle$, for some graph $H$ of treewidth at most $t$, then $G$ has an $H'$-partition of layered width 1 with respect to the same layering, for some graph $H'$ of treewidth at most $(t+1)\ell-1$.  That is, if $G\subseteq H\boxtimes P\boxtimes K_\ell$ for some graph $H$ of treewidth at most $t$  and for some path $P$, then $G\subseteq H' \boxtimes P$ for some graph $H'$ of treewidth at most $(t+1)\ell-1$.
\end{lem}

@@

Using relationship between $k$-planar graphs and $(k+1,2)$-shortcut systems along with a direct application of \cref{modern-major-general} we can obtain a slightly weaker version of \cref{k-planar}, below. The minor modifications needed to obtain the stronger bound are described in \cref{sec-k-planar}.

 % of planar graphs shows that \cref{modern-major-general} has immediate applications to $k$-planar graphs.  The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

@@

% \cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $O(k^3)$
%  % $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$
% on the layered width. 
% % \note{DW}{I would just write, ``\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width.}
% We prove the stronger bound in \cref{sec-k-planar}. 
% %Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% % PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to a subgraph of $G^\PP$ for some graph $G$ of Euler genus at most $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

Finally, we state the following corollary of \cref{ApexMinorFree,modern-major-general,MakeWidth1}. This is the most general result that follows from \cref{modern-major-general} and that of \citet{dujmovic.joret.ea:planar}.

@@

%\note{DW}{should this subsection be moved to the end of \cref{Tools}?}\note{PM}{No, I don't think so. The information given in Section 2 isn't detailed enough to compare with layered treewidth versus layered partitions.}

@@

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, suppose that an $n$-vertex graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$. \citet{dujmovic.joret.ea:planar} proved that $G$ has layered treewidth at most $(k+1)\ell$. 
% So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. 
What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. 
The \emph{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For any node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

We begin with a fairly standard lemma about normalized tree decompositions.

\begin{lem}\lemlabel{nice-decomposition}
  For every graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

We are now ready to prove our main result, which we restate here convenience:

@@

  Thus, $a_0,\ldots,a_p$ is a lazy walk in $H^+$ of length $p\le k$ where the distance $s_i$ between $a_i$ and the root $x_0$ of $T$ is non-decreasing.  By removing repeated vertices this gives a path in the directed graph $\overrightarrow{H}^+$ obtained by directing each edge $xy\in E(H^+)$ from its $T$-descendant $x$ towards its $T$-ancestor $y$. 
  Finally, we are in a position to appeal to \cite[Lemma~24]{pilipczuk.siebertz:polynomial-arxiv} which states that the number of nodes in $\overrightarrow{H}^+$ that can be reached from any node $x$ by a directed path of length at most $k$ is at most $\binom{k+t}{t}$.
\end{proof}

@@

This section shows that the constants in \cref{modern-major-general} can be improved in the case of $k$-planar graphs. Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is not critical since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap V_j$ for $j > k+1$.
% \end{compactitem}
% \note{DW}{do we need the bulletted list here?}
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| &\le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)\\
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

@@

  % Note that $M$ is a Mercedes graph 
  % \note{DW}{``Mercedes'' not defined} and that each of the faces

@@

  The proof finishes in the same way as the proof in \cite{dujmovic.joret.ea:planar}.  The paths $P_1,\ldots,P_k$, and $S=V(Y^+)$ become elements of the $H$-partition. 
  % \note{DW}{Do we need $S$ here?}
   Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}, except that now each node $x$ has a child for each face $F_i$ of $M^+_x$ that contains a vertex of $G$ in its interior.

@@

\begin{proof}[Proof of \cref{1-planar}]
Given a 1-plane graph $G$, add edges to make it edge maximal so that it has an outer face $F=v_1v_2v_3$. Next, add a vertex $r$ adjacent to $v_1$, $v_2$, and $v_3$ to obtain an edge-maximal 1-plane graph $\overline{G}$ with one vertex $r$ of degree 3 on its outer face.

@@

% \noindent\note{PM}{Double-check the calculation for \cref{k-planar} and \cref{g-k-planar} and then fix the constants here.}\note{DW}{done}

@@

\begin{lem}[\citep{pilipczuk.siebertz:polynomial-soda}]
\label{p-centered-treewidth}
Every graph $H$ of treewidth at most $t$ has $\chi_p(H)\in (t+1)\binom{p+t}{t}$.
\end{lem}

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
%\usepackage{iwona}
\usepackage{thm-restate}

@@

%\usepackage[mathlines]{lineno}
%% \setlength{\linenumbersep}{2em}
%\linenumbers
%% \rightlinenumbers
%% \linenumbers
%\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
% \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
% \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
% \renewenvironment{#1}%
%    {\linenomath\csname old#1\endcsname}%
%    {\csname oldend#1\endcsname\endlinenomath}}% 
%\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
% \patchAmsMathEnvironmentForLineno{#1}%
% \patchAmsMathEnvironmentForLineno{#1*}}%
%\AtBeginDocument{%
%\patchBothAmsMathEnvironmentsForLineno{equation}%
%\patchBothAmsMathEnvironmentsForLineno{align}%
%\patchBothAmsMathEnvironmentsForLineno{flalign}%
%\patchBothAmsMathEnvironmentsForLineno{alignat}%
%\patchBothAmsMathEnvironmentsForLineno{gather}%
%\patchBothAmsMathEnvironmentsForLineno{multline}%
%}
%
%
%

@@

\setlength{\parskip}{0.85ex}

 \newcommand{\note}[2]{{\color{red}[#1:~#2]}}
% \newcommand{\notex}[2]{{\color{blue}[#1:~#2]}}
\newcommand{\notex}[2]{}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992);  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002); and \citet{bonamy.gavoille.ea:shorter} use it to find shorter implicit representations of planar graphs (making progress on a sequence of results going back to 1988 \citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}).

@@

Although $k$-planar graphs are the most high-profile target for a generalization of \cref{PlanarBasic}, we actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

This definition is related to $k$-planarity because of the following observation:

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is a subgraph of $G^\PP$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centered colouring  \citep{micek:personal}, 
% and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

%\notex{PM}{TODO: Write here about applications. non-repetitive colouring, $p$-centered colouring, $(g,k)$-planar graph, $(g,d)$-map graphs, $k$-NN graphs, David's rough characterization.}

@@

We now formally define $k$-planar graphs.  An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$.  Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

% \notex{DW}{I  prefer to not define ``layered $H$-partition''. Just define the ``layered width'' of an $H$-partition, which is what we did in \citep{dujmovic.joret.ea:planar}. There are many places where we write ``$H$-partition with layered width'' without writing ``layered $H$-partition''. I prefer it this way.}

@@

\begin{lem}[\citep{dujmovic.joret.ea:planar}] 
\label{MakeWidth1}
If a graph $G$ has an $H$-partition of layered width $\ell$  
%with respect to layering $\langle V_0,V_1,\dots\rangle$, 
for some graph $H$ of treewidth at most $t$, then $G$ has an $H'$-partition of layered width 1   
%with respect to the same layering, 
for some graph $H'$ of treewidth at most $(t+1)\ell-1$.  That is, if $G\subseteq H\boxtimes P\boxtimes K_\ell$ for some graph $H$ of treewidth at most $t$  and for some path $P$, then $G\subseteq H' \boxtimes P$ for some graph $H'$ of treewidth at most $(t+1)\ell-1$.
\end{lem}

@@

Using the relationship between $k$-planar graphs and $(k+1,2)$-shortcut systems along with a direct application of \cref{modern-major-general} we can obtain a slightly weaker version of \cref{k-planar}, below. The minor modifications needed to obtain the stronger bound are described in \cref{sec-k-planar}.

@@

% \cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $O(k^3)$
%  % $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$
% on the layered width. 
% % \notex{DW}{I would just write, ``\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width.}
% We prove the stronger bound in \cref{sec-k-planar}. 
% %Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% % PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is a subgraph of $G^\PP$ for some graph $G$ of Euler genus at most $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

Finally, we state the following corollary of \cref{ApexMinorFree,modern-major-general,MakeWidth1}. This is the most general result that follows from \cref{modern-major-general} and the work of \citet{dujmovic.joret.ea:planar}.

@@

%\notex{DW}{should this subsection be moved to the end of \cref{Tools}?}\notex{PM}{No, I don't think so. The information given in Section 2 isn't detailed enough to compare with layered treewidth versus layered partitions.}

@@

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, 
\citet{dujmovic.joret.ea:planar} proved that if a graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$, then $G$ has layered treewidth at most $(k+1)\ell$. 
% So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. 
What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

@@

For a tree $T$ rooted at some node $x_0\in V(T)$, we we say that a node $a\in V(T)$ is a \emph{$T$-ancestor} of $x\in V(T)$ (and $x$ is a \emph{$T$-descendant} of $a$) if $a$ is a vertex of the path, in $T$, from $x_0$ to $x$.  Note that each node $x\in v(T)$ is a $T$-ancestor and $T$-descendant of itself.  We say that a $T$-ancestor $a\in V(T)$ of $x\in V(T)$ is a \emph{strict} $T$-ancestor of $x$ if $a\neq x$. 
The \emph{$T$-depth} of a node $x\in V(T)$ is the length of the path, in $T$, from $x_0$ to $x$.  For each node $x\in V(T)$, define
\[T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}] \]
to be the maximal subtree of $T$ rooted at $x$.

We begin with a fairly standard lemma about normalised tree decompositions.

\begin{lem}\lemlabel{nice-decomposition}
  For every graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ with width $t$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each node $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

@@

We are now ready to prove our main result, which we restate here for convenience:

@@

  Thus, $a_0,\ldots,a_p$ is a lazy walk in $H^+$ of length $p\le k$ where the distance $s_i$ between $a_i$ and the root $x_0$ of $T$ is non-decreasing.  By removing repeated vertices this gives a path in the directed graph $\overrightarrow{H}^+$ obtained by directing each edge $xy\in E(H^+)$ from its $T$-descendant $x$ towards its $T$-ancestor $y$. 
  Finally, we are in a position to appeal to \cite[Lemma~24]{pilipczuk.siebertz:polynomial-soda} which states that the number of nodes in $\overrightarrow{H}^+$ that can be reached from any node $x$ by a directed path of length at most $k$ is at most $\binom{k+t}{t}$.
\end{proof}

@@

This section shows that the constants in \cref{modern-major-general} can be improved in the case of $k$-planar graphs. Let $G$ be a $k$-plane graph.  We will assume, for ease of exposition, that any point $p\in\R^2$ is involved in at most one crossing $(p,vw,xy)$ of $G$. This assumption is justified since it can be enforced by a slight deformation of the edges of $G$ and the resulting (deformed) graph is also $k$-plane.

@@

We can reduce the layered width of the $H$-partition of $G$ from $O(k^3)$ to $O(k^2)$ by observing that the dummy vertices in $V(G_0)\setminus V(G)$ do not contribute to the layered width of this partition.  In this setting, the proof of \cref{clm:general-width} is simpler since each vertex $w\in Y_x$ contributes at most two vertices to $V_i\cap Y_x$.  More precisely, each path $P\in\mathcal{P}$ containing an internal (dummy) vertex $w\in Y_x\cap (V_{i-j}\cup V_{i+j})$ contributes: (i)~at most two vertices to $S_x\cap V_i$ for $j\in\{0,\ldots,\floor{(k+1)/2}\}$; (ii)~at most one vertex to $S_x\cap V_j$ for $j\in\{\floor{(k+1)/2},\ldots,k+1\}$; or (iii)~no vertices to $S_x\cap V_j$ for $j > k+1$.
Redoing the calculation at the end of the proof of \cref{clm:general-width} then yields 
\begin{align*}
    |S_x\cap Y_i| \le d\ell\left( 
       2 
       + 4\left\lfloor\frac{k+1}{2}\right\rfloor 
       + 2\left(k+1-\left\lfloor\frac{k+1}{2}\right\rfloor\right)
       \right)
       & =
       d\ell\left(
         2k + 2\left\lfloor\frac{k+1}{2}\right\rfloor + 4
       \right) \\
       & \le
       d\ell(3k+5) 
      = 18k+30 \enspace .
\end{align*}    
With this change, the layered width of the partition given by \cref{modern-major-general} becomes $18k^2+30k$.  This establishes \cref{k-planar}.

@@

  % Note that $M$ is a Mercedes graph

@@

  The proof finishes in the same way as the proof in \cite{dujmovic.joret.ea:planar}.  The paths $P_1,\ldots,P_k$, and $S=V(Y^+)$ become elements of the $H$-partition. 
   Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}, except that now each node $x$ has a child for each face $F_i$ of $M^+_x$ that contains a vertex of $G$ in its interior.

@@

\begin{proof}[Proof of \cref{1-planar}]
Given a 1-plane graph $G$, add edges to make it edge-maximal so that it has an outer face $F=v_1v_2v_3$. Next, add a vertex $r$ adjacent to $v_1$, $v_2$, and $v_3$ to obtain an edge-maximal 1-plane graph $\overline{G}$ with one vertex $r$ of degree 3 on its outer face.

@@

@@

\begin{lem}[\citep{pilipczuk.siebertz:polynomial-soda}]
\label{p-centered-treewidth}
Every graph $H$ of treewidth at most $t$ has $\chi_p(H)\leq  (t+1)\binom{p+t}{t}$.
\end{lem}

The following lemma is implicitly due to \citet{micek:personal}. We include the proof for completeness.

% \note{DW}{can we replace by a result of \citet{micek:personal}?}
% \note{PM}{I didn't see a statement of this exact result in the last draft Piotr sent.  Though their Theorem~7 is this proof for the special case of planar graphs.}
